---
title: "Gaussian Mixture Models study"
author: "Anna Kubasiak"
date: "13 March 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(magrittr)
library(mcsm)
library(mvtnorm)
library(ellipse)
library(ggplot2)
# library(knitr)
```
This is the first chapter of Mixture Model Tutorial with R. In this chapter we will start with the simplest case of Gaussian Mixture Model where the average $\mu$ and covariance matrix $\Sigma$ of each component model is known as well as the proportion of each component $\pi$ and the number of components $K$. Our goal in this chapter will be to find out from which component distribution those points come from: $p(x_i)$

## Gaussian mixture 
In this example we will work with 2D gaussian distributions. We set:

```{r generate distributions, echo=FALSE, message = FALSE}
set.seed(13)

#Number of distributions
K=3
d=2
s<-0.6
nSim<-1000
maxMu<-4
# parameters of each distribution
mu<-list()
sigma<-list()
ellip<-list()
for (i in 1:K){
  mu[[i]]<-sample(1:maxMu,d)
  mat1<-matrix(rnorm(d^2,0,s),ncol=d)
  sigma[[i]]<-mat1%*%t(mat1) %>% round(2)
  ellip[[i]]<-ellipse(sigma[[i]],centre=mu[[i]])
}
# color for each distribution
distColor<-data.frame(hsv=hsv((1:K)/K))
distColor$distR=distColor$hsv %>% substr(2,3) %>% as.hexmode %>% as.integer
distColor$distG=distColor$hsv %>% substr(4,5)%>% as.hexmode %>% as.integer
distColor$distB=distColor$hsv %>% substr(6,7)%>% as.hexmode %>% as.integer


# prior probabilities 
temp<-runif(K,min=0,max=1)
pi<-temp/sum(temp)
muDf<-as.data.frame(mu)
 names(muDf)<-paste0("mu",1:K)
 
```
K=`r I(K)`

```{r,fig.width=3, fig.height=3, distributions, message=FALSE, echo=FALSE}

p<-ggplot()+ 
  theme_bw()+
  ggtitle(label='Contours of 2d gaussians')
  
 
for (k in 1:K){
  
  muLabel<-paste(expression(mu),"=","(",(paste(mu[[k]],collapse=',')),")",sep='')
  p<-p+
    geom_path(data=as.data.frame(ellip[[k]]),aes(x,y),color=distColor$hsv[k],lwd=1.5)
    # geom_point(aes(x=mu[[k]][1],y=mu[[k]][2]),lwd=2,color=distColor$hsv[k])
    
}
p+geom_point(aes(x=as.numeric(muDf[1,]),y=as.numeric(muDf[2,])),lwd=2,color=distColor$hsv)
    
# p+geom_point(data=allX, aes(x=d1,y=d2),alpha=0.5,size=0.9)
# 
# 
# ggplot(allX, aes(x = d1, y = d2)) + geom_point() +geom_density2d()+
#     geom_path(data=as.data.frame(ellip[[k]]),aes(x,y),color=distColor$hsv[k])

```


